Reversible Higher-Order Programs \\ (Functional Pearl)

### 2013-03-27 rc13/frac2.tex

\newtheorem{definition}{Definition}

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include Pi
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! [[ = \llbracket
%! ]] = \rrbracket
%! ^^^ = ^{\dagger}
%! eta* = \eta
%! eps* = \epsilon
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! -> = \rightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! Sigma = \Sigma
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! Pow = \mathcal{P}
%! Pi = \Pi
%! PiT = \Pi^{o}
%! PiEE = \Pi^{\eta\epsilon}_{+}
%! PiT = \Pi^{o}
%! PiTF = \Pi^{/}
%! bullet = \bullet
%! * = \times

@@

\title{Fractional Types: \\
First-Class Pattern Matching Clauses and Higher-Order Reversible Functions}
\authorinfo{Jacques Carette \and Amr Sabry}
           {McMaster University \and Indiana University}
           {carette@mcmaster.ca \and sabry@indiana.edu}

@@

We are witnessing a convergence of ideas from several distinct research
communities (physics, mathematics, and computer science) towards replacing
\emph{equalities} by \emph{isomorphisms}. The combined programme has sparked
a significant amount of research that unveiled new and surprising connections
between geometry, algebra, logic, and computation (see~\cite{baez2011physics}
for an overview of some of the connections).

In the physics community, Landauer~\cite{Landauer:1961,Landauer},
Feynman~\cite{springerlink:10.1007/BF02650179}, and others have interpreted
the laws of physics as fundamentally related to computation. The great
majority of these laws are formulated as equalities between different
physical observables which is unsatisfying: \emph{different} physical
observables should not be related by an \emph{equality}. It is more
appropriate to relate them by an \emph{isomorphism} that witnesses, explains,
and models the process of transforming one observable to the other.

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive
mathematics~\cite{citeulike:7374951}. This theory has been further extended
with \emph{identity types} representing the proposition that two terms are
``equal.'' (See~\cite{streicher,warren} for a survey.) Briefly speaking,
given two terms $a$ and $b$ of the same type $A$, one forms the type
$\texttt{Id}_A(a,b)$ representing the proposition that~$a$ and~$b$ are equal:
in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
models the process of transforming $a$ to $b$ and vice-versa.

%% AMR: I am not sure how to take this comment into account. I think we
%% should make a specific connection to Martin-Lof type theory to at least
%% justify the mention of groupoids in the conclusion. I agree though that
%% this is a very technical point and I do find it hard to express the point
%% in a couple of sentences. Suggestions welcome.

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated by the pragmatic concern of searching large libraries of functions
by providing one of the many possible isomorphic types for the desired
function~\cite{Rittri:1989:UTS:99370.99384}. More recently, type isomorphisms
have taken a more central role as \emph{the} fundamental computational
mechanism from which more conventional, i.e., irreversible computation, is
derived. In our own previous
work~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012} we started with the
notion of type isomorphism and developed from it a family of programming
languages, {{Pi}} with various superscripts, in which computation is an
isomorphism preserving the information-theoretic entropy.
%% Moreover these languages have models in various flavors of symmetric
%% monoidal categories (which are the backbone of models for linear logic and
%% quantum computation).
%% We have also invested a significant amount of time in using {{Pi}} as
%% an actual programming language and established several idioms for
%% programming large classes of interesting (recursive) programs in
%% {{Pi}}.

A major open problem remains, however: a higher-order extension of
{{Pi}}. This extension is of fundamental importance in all the originating
research areas.
%% In physics, it allows a process or observer to be treated as
%% ``data'' that can be transformed by higher-level processes or observed by
%% meta-level observers. 
In physics, it allows for quantum states to be viewed as processes and
processes to be viewed as states, such as with the Choi-Jamiolkowski
isomorphism~\cite{choi1975completely,jamiolkowski1972linear}.  In mathematics
and logic, it allows the equivalence between different proofs of type
$\texttt{Id}_A(a,b)$ to itself be expressed as an isomorphism (of a higher
type) $\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code.

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{closed category} from the underlying monoidal
category for~{{Pi}}. Although the general idea of such a construction is
well-understood, the details of adapting it to an actual programming language
are subtle.  Our main novel technical device to achieving the higher-order
extension is a \emph{fractional type} which represents \emph{negative
  information} and which is so named because of its duality with conventional
product types.  
%% By adding a formal (linear) dual to (type theoretic) products, we can
%% internalize the notion that ``taking an input'' corresponds
%% to negative information, which turns out to be the basis of first-class
%% functions.  
The remainder of the paper reviews {{Pi}} and then introduces the syntax and
semantics of the extension with fractional types. We then study the
properties of the extended language and establish its expressiveness via
several constructions and examples that exploit its ability to model
higher-order computations.

%% AMR:
%% I think the previous paragraph in the intro that talks about Martin-Lof type 
%% theory provides the context. Anything else??

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background: {{Pi}} }
\label{sec:pi}

@@

\noindent 
Types include the unit type {{1}}, sum types {{b1+b2}}, and
product types {{b1*b2}}.  Values include {{()}} which is the only value of
type {{1}}, {{left v}} and {{right v}} which inject~{{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. 
%% But these values should be regarded as largely ancillary: we do not
%% treat them as first-class citizens, and they only occur when observing
%% the effect of an isomorphism.

%% \paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
%% (value) type isomorphisms.  They are also typed, by the shape of the (value)
%% type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
%% the following isomorphisms:
\paragraph*{Isomorphisms.} The terms of {{Pi}} witness
type isomorphisms of the form {{b <-> b}}. They consist of base
isomorphisms, as defined below, and their composition.
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}r c l@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 & <-> & b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) & <-> & (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b & <-> & b &: identr*
%swap* :&  b1 * b2 & <-> & b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) & <-> & (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 & <-> & (b1 * b3) + (b2 * b3)~ &: factor

@@

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}: 
%subcode{bnf} include main
% top level term, l ::= c v

% \noindent
% To summarize, the syntax of {{Pi}} is given as follows.

% \begin{definition}{(Syntax of {{Pi}})}
% \label{def:Pi}
% %subcode{bnf} include main
% % value types, b ::= 1 | b+b | b*b 
% % values, v ::= () | left v | right v | (v,v) 
% %
% % iso.~types, t ::= b <-> b
% % base iso ::= swap+ | assocl+ | assocr+ 
% %     &|& unite | uniti | swap* | assocl* | assocr* 
% %     &|& dist | factor 
% % iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% % top level term, l ::= c v
% \end{definition}

@@

% Operationally, the
% semantics of {{Pi}} is given using two mutually recursive interpreters: one
% going forward and one going backwards. The use of {{sym}} switches control
% from one evaluator to the other. We will present the operational semantics in
% the next section along with the extension with fractional types and
% values. For now, we state without proof that the evaluation of well-typed
% combinators always terminates and that {{Pi}} is logically reversible, i.e.,
% that for all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 :
%     b2}} we have the forward evaluation of {{c v1}} produces {{v2}} iff the
% backwards evaluation of {{c v2}} produces {{v1}}.

%% AMR:
%% \jacques{Since we have Agda code for the above, should we refer to it
%% here, and make it available?}
%% That is a great idea but it would take a significant amount of time to
%% clean up the code and make it presentable. If someone is willing to do
%% that, then we should definitely add a reference to the code and make it
%% available.
%% Jacques:  Right.  I'll definitely work on cleaning up the paper first,
%% then see if there is time left for the code.

%% As is usual with type theories,
%% the coherence conditions (pentagonal and triangle identity) are not explicit,
%% but would correspond to certain identity types being trivial.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Language: {{PiTF}} }

The language {{Pi}} models isomorphisms of values rather well, and as we
established before~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012}, it is
logically reversible and each computation it expresses preserves the
information-theoretic entropy. However, purposefully, it has no distinguished
notion of \emph{input} or \emph{output}; more precisely, because of strict
preservation of information, these two concepts coincide.  This is the root
of reversibility. But if we want to model functions of any flavor as values,
we need to differentiate between these notions. Our idea, inspired by
computational dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
polarity~\cite{Girard87tcs,10.1109/LICS.2010.23}, and the categorical notion
of \emph{compact
  closure}~\cite{Selinger:2007:DCC:1229185.1229207,Abramsky:2004:CSQ:1018438.1021878},
is to introduce a \emph{formal dual} for our types.  We thus consider a type
in a negative position to possess negative information, as it is really a
request for information.  Since information is logarithmic, this means that a
request for information should behave like a \emph{fractional type}.

% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
% Jacques: Exactly!  1/a is negative information is input, b is positive is
%    output.

% AMR: relations are indeed only presented as a way to explain the semantics

The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
simple:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v

\noindent For a given type {{b}}, the values of type {{1/b}} are of the form
{{1/v}} where {{v : b}}. Note that {{1/v}} is purely formal, as is
{{1/b}}.

Semantically, we are extending the symmetric monoidal category modeling
{{Pi}} to a \emph{compact closed} one, i.e., a category in which morphisms
are representable as objects. In such a setting, the new dual objects (i.e.,
the fractionals) must satisfy the following isomorphism witnessed by two new
type indexed combinators {{eta*_b}} and {{eps*_b}}:
\begin{multicols}{3}
  \dgm{eta_times2}

\columnbreak

%subcode{bnf} include main
%! columnStyle = r r c l l
%eta*_b :&  1 & <-> & 1/b * b &: eps*_b
%% XXX: diagrams need unit wires(?)
%% Roshan: We should just be clear that unit wires are dropped. Dgms
%% including them dont communicate any additional information and
%% appear crowded.

\columnbreak

  \dgm{eps_times2}  
\end{multicols}

\vspace{-10pt}
\noindent From a programming perspective, we think of a value {{1/v}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{v}}. In other words, {{1/v}} is a \emph{pattern} representing the absence of
some information of a specific shape (i.e., \emph{negative information}),
that can only be reconciliated by an actual value {{v}}. The combinator
{{eta*_b : 1 <-> 1/b * b}} thus represents a fission point which creates ---
out of no information --- an equal amount of negative and positive
information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
up an equal amount of negative and positive information producing no residual
information.

Historically, versions of compact closed categories were introduced by
Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the generalization of
monoidal categories to model quantum computation.  The first and most
intuitive example of such categories is the category of finite sets and
relations. In that case, the dual operation (i.e., the fractional type)
collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
interpretation of fractionals, this is still an interesting category as it
provides a model for a reversible higher-order language. Indeed, any relation
is reversible via the standard \emph{converse} operation on a relation.
Furthermore, any relation between {{b1}} and {{b2}} can be represented as 
a subset of {{b1 * b2}}, i.e., as a value.

In this section, we provide a simple semantics of {{PiTF}} in the
category of finite sets and relations, in the following way: We
interpret each combinator {{c : b1 <-> b2}} as a \emph{relation}
between the sets denoted by {{b1}} and {{b2}} respectively. Sets
{{b1}} and {{b2}} can be related only if the same information measure.
As we illustrate, this semantics allows \emph{arbitrary} relations
(including the empty relation and relations that are not isomorphisms)
to be represented as values in {{PiTF}}.  In Sec.~\ref{sec:cat} we
discuss more refined semantics that are more suitable for richer
categories.

\begin{definition}[Denotation of Value Types]
\label{chx:def:denot}
Each type denotes a finite set of values as follows:
%subcode{opsem} include main
%! <- = \leftarrow
%! union = \cup
% [[1]] '= {[ () ]}
% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
% [[1/b]] '= {[ 1/v ~|~ v <- [[b]] ]}
\end{definition}

We specify
relations using a deductive system whose judgments are of the form 
{{ v1 ~~c~~ v2 }} indicating that the pair {{(v1,v2)}} is in the relation 
denoted by {{c}}.

\begin{definition}[Relational semantics]
\label{def:relational-PiTF}
Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
as specified below.~\footnote{
In the interest of brevity, we have treated the type annotations in the base
isomorphisms as implicit; for example, the rule for {{assocr*}} should really
read:

%subcode{proof} include main
%@ v1 in b1
%@ v2 in b2
%@ v3 in b3
%@@ ((v1, v2), v3) ~~assocr*_{b1, b2, b3}~~ (v1, (v2, v3))
}
%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%----
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%----
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%----
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%
%@ ~
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()
\end{definition}

The semantics above defines a relation rather than a function. It is still
reversible in the sense that it ``mode-checks''. In other words, for any
combinator~{{c}} we can treat either its left or its right side as an input;
this allows us to define {{sym}} as just relational converse, which amounts
to an argument swap. We can then give an operational interpretation to the
semantic relation by defining two interpreters: a forward evaluator and a
backwards evaluator that exercise the relation in opposite directions, but
which now can return a \emph{set} of results. This operational view exploits
the usual conversion of a relation from a subset of {{A * B}} to a function
from {{A}} to the powerset of {{B}} whose composition is given by the Kleisli
composition in the powerset monad. It is straightforward to check that if the
forward evaluation of {{c v1}} produces {{v2}} as a possible answer then the
backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.  One
can check that all relations for {{Pi}} are total functional relations whose
converse are also total functional relations -- aka isomorphisms.  However
{{eta*_b}} relates {{()}} to multiple values (one for each value of type {{b}}),
and is thus not functional; {{eps*_b}} is exactly its relational converse, as
expected.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Expressiveness and Examples}

Having introduced the syntax of our extended language, its semantics in the
category of sets and relations, and its operational semantics using forward
and backwards interpreters, we now illustrate its expressiveness as a
higher-order programming language. In the following presentation consisting
of numerous programming examples, we use {{bool}} as an abbreviation of
{{1+1}} with {{true}} as an abbreviation for {{left ()}} and {{false}} as an
abbreviation for {{right ()}}. In addition, instead of presenting the code
using the syntax of {{Pi}}, we use circuit diagrams that are hopefully more
intuitive. Each diagrams represents a combinator whose evaluation consists of
propagating values along the wires.  For the sake of readability, we omit
obvious re-shuffling circuitry and elide trivial unit wires which carry no
information.\footnote{The full code is available at
\url{www.cas.mcmaster.ca/~carette/PiFractional}.}

%%%%%%%%%%%%
\subsection{First-Class Relations}

The most basic additional expressiveness of {{PiTF}} over {{Pi}} is the
ability to express relations as values. Indeed a value of type {{1/b1 * b2}}
is a pair of a \emph{constraint} that can only be satisfied by some 
{{v1 : b1}} and a value {{v2 : b2}}. In other words, it corresponds to a
function or relation which when ``given'' a value {{v1 : b1}} ``releases''
the value {{v2 : b2}}. To emphasize this view, we introduce the abbreviation:

%subcode{bnf} include main
%  b1 :-* b2 &::=& 1/b1 * b2

\noindent which suggests a function-like behavior for the pair of a fractional value
and a regular value. What is remarkable is that we can almost trivially turn
any combinator {{c : b1 <-> b2}} into a (constant) value of 
type {{b1 :-* b2}} as shown below on the left:

%subcode{opsem} include main
%! columnStyle = rcl
% name &:& (b1 <-> b2) -> (1 <-> (b1 :-* b2))
% name c &=& eta*_{b1} (;) (id (*) c)

\begin{multicols}{2}
  \dgm{function-times}

  \dgm{delimc-times}
\end{multicols}
\noindent Dually, as illustrated above on the right, we also have a
{{coname}} combinator which when given an answer {{v2 : b2}}, and a request
for the corresponding input {{1/b1}} for {{c : b1 <-> b2}}, eliminates this
as a un-needed computation (see also Sec. 3, \cite{abramsky-2008}).

More generally any combinator manipulating values of type {{b}} can be turned
into a combinator manipulating values of type {{1/b}} and vice-versa. This is
due to the fact that fractionals types satisfy a self-dual involution
relating {{b}} and {{1/(1/b)}}:

%subcode{opsem} include main
%! columnStyle = rcl
% doubleDiv &:& b <-> 1/(1/b)
% doubleDiv &=& uniti (;) (eta*_{1/b} (*) id) (;) assocr* (;) (id (*) eps*_b) (;) swap* (;) unite

% jacques: elide for now, to save space, as this diagram is backwards.
% \dgm{involution-times}

% As a simple example, consider {{swap+}} denoting boolean
% negation.  It is straightforward to check that the combinator 
% {{name~swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:

% %subcode{proof} include main
% %@ ~
% %@@ () ~~(name~swap+)~~ (1/false,true)
% %
% %@ ~
% %@@ () ~~(name~swap+)~~ (1/true,false)

% \noindent The relation {{name~swap+}} maps {{()}} to a value which can be used
% to convert any {{false}} to {{true}} and vice-versa. We formalize this
% ``apply'' combinator below.

%%%%%%%%%%%%
\subsection{Higher-Order Relations}

We are now a small step from implementing various higher-order combinators
that manipulate functions or relations. In particular, we can \emph{apply},
\emph{compose}, \emph{curry}, and \emph{uncurry} values representing
relations.  We first show the realization of \emph{apply}:

%subcode{opsem} include main
%! columnStyle = rcl
% apply &:& (b1 :-* b2) * b1 <-> b2
% apply &=& swap* (;) assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite

\begin{multicols}{2}
  \dgm[1.2]{apply-times1}

  \dgm{apply-times2}
\end{multicols}

\noindent Intuitively, we simply match the incoming argument of type {{b1}} with
the constraint encoded by the function. If they match, they cancel
each other and the value of type {{b2}} is exposed with no
constraints. Otherwise, the result is undefined.

\noindent A flipped variant is also useful:
%subcode{opsem} include main
%! columnStyle = rcl
% apply' &:& b1 * (b1 :-* b2) <-> b2
% apply' &=& assocl* (;) (swap* (*) id) (;) (eps*_{b1} (*) id) (;) unite

\noindent Function or relation composition is now straightforward:

%subcode{opsem} include main
%! columnStyle = rcl
% compose &:& (b1 :-* b2) * (b2 :-* b3) -> (b1 :-* b3)
% compose &=& assocr* (;) (id (*) apply')

\begin{multicols}{2}
  \dgm[1.0]{compose-times1}

  \dgm{compose-times2}
\end{multicols}

% \dgm{compose-times3}

\noindent
We can also derive currying (and dually, uncurrying) combinators. Observe that
the type of {{curry}} needs to be {{(b1 * b2 :-* b3) <-> (b1 :-* (b2 :-* b3))}},
which can be written in mathematical notation as 
{{1/(b1 * b2) * b3 = (1/b1) * ((1/b1) * b3)}}. This means {{curry}} can be
written using {{recip}}, the implementation of the mathematical identity
{{1/(b1 * b2) = 1/b1 * 1/b2}}:

%subcode{opsem} include main
%! columnStyle = rcl
% recip &:& 1 / (b1 * b2) <-> 1 / b1 * 1 / b2
% recip &=& (uniti) (;) (uniti) (;) (assocl*) (;) ((eta*_{b1} (*) eta*_{b2}) (*) id) (;)
%     && (reorder (*) id) (;) (assocr*) (;) (id (*) swap*) (id (*) eps*) (;) swap* (;) (unite)

\noindent
where {{reorder}} is the obvious combinator of type {{b1 * (b2 * b3) * b4 <-> b1 * (b3 * b2) * b4}}.
With {{recip}} out of the way, we can easily write {{curry}}:

%subcode{opsem} include main
%! columnStyle = rcl
% curry &:& b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)
% curry &=& (recip * id) (;) assocr*

\noindent That {{recip}} is the heart of currying seems quite remarkable.

%%%%%%%%%%%%
\subsection{Feedback, Iteration, and Trace Operators}

Mathematically speaking, recursion and iteration can be expressed using
categorical trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
language like {{Pi}} there are two natural families of trace operators that
can be defined, an additive family (explored in detail in our previous
work~\cite{rc2011}) and a multiplicative family, which is expressible using
fractionals.

The idea of the multiplicative trace operator is as follows. We are given a
computation {{c : b1 * b2 <-> b1 * b3}}, and we build a ``looping'' version
which feeds the output value of type {{b1}} back as an input. Effectively,
this construction cancels the common type {{b1}}, to produce a new combinator
{{trace*~c:b2<->b3}}.

With fractionals, {{trace*}} becomes directly expressible:

%subcode{opsem} include main
%! columnStyle = rcl
% trace*_b &:& ((b * b1) <-> (b * b2)) -> (b1 <-> b2)
% trace*_b c &=& uniti (;) (eta*_b (*) id) (;) assocr* (;) 
%            && (id (*) c) (;) assocl* (;) (eps*_b (*) id) (;) unite

As an example, we can use the operational semantics (outlined in the previous
section) to calculate the result of applying {{trace*_{bool} (swap+ (*) id)}}
to {{false}}.
%subcode{opsem} include main
%! columnStyle = ll
% {[ ((),false) ]}  & (uniti)
% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool} * id)
% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (id (*) not )
% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
% emptyset & (eps*_{bool} * id)

\noindent The computation with {{true}} gives the same result. This confirms
that although our semantics is formally reversible, it does not result in
isomorphisms.

More abstractly, the evaluation of {{trace*~c : b2 <-> b3}} must ``guess'' a
value of type {{b1}} to be provided to to the inner combinator 
{{c : b1 * b2 <-> b1 * b3}}. 
This value {{v1 : b1}} cannot be arbitrary: it must be such that it is the value
produced as the first component of the result.  In general, there may be
several such fixed-point values, or none. Indeed with a little ingenuity 
(see the next section), one can 
express any desired relation by devising a circuit which keeps the desired pairs as 
valid fixed-points.

% \begin{example}
% Consider the combinator {{c : bool * 1 <-> bool * 1}}, with {{c = id}}.

% \noindent Using {{trace*}}, we construct the
% combinator {{trace* c : 1 <-> 1}}. Applying this to 
% {{() : 1}} requires us to find a value {{b : bool}} such that 
% {{(b,()) ~~d~~ (b,())}}. Given that the type {{bool}} 
% has two values, there are two values {{false}} and {{true}} that satisfy the 
% constraint. In other words, {{() ~~(trace* c)~~() }}.
% \end{example}

% \begin{example}
% \label{ch3:ex:annihilate}
% Consider a small variation on the combinator {{c}} above:

% {{c : bool * 1 <-> bool * 1}}

% {{c = swap+ (*) id }}

% \noindent which negates the boolean component of its incoming
% pair. Using the multiplicative trace operator, we can construct the
% combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
% combinator to {{() : 1}} requires us to find a value {{b : bool}} such
% that {{(b,()) ~~swap+~~ (b,())}} which is impossible since {{swap+}} 
% has no fixed points.  Operationally, the evaluation of
% such a combinator should produce no value (were it to even terminate).
% \end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{(Finite) Relational Programming}
\label{ch3:sec:lp}

Relational programming leverages set-theoretic relations and their
composition to express computational tasks in a declarative way. 
% For instance, consider this example:
% {{parent = {[ (A,B), (B,C), (A,D) ]} }}
% {{grandparent = parent (o) parent}}
% \noindent The example defines a relation {{parent}} specified using a
% set of tuples and another relation {{grandparent}} specified using the
% composition of two parent relations. If we wrote this example in a
% relational language (e.g., Prolog) and we executed the query
% {{grandparent(A)}}, we would get the answer {{ {[C]} }}.
It turns out that with the addition of the multiplicative trace, and
the move to relations motivated in the previous section, we can
express relational programming.

Consider the relation {{R}} on booleans given by:

>>>>

\newtheorem{definition}{Definition}
\newcommand{\bra}[1]{\langle #1 |}
\newcommand{\singleton}[1]{\textsc{Sing}~#1}

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include Pi
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! [[ = \llbracket
%! ]] = \rrbracket
%! ^^^ = ^{\dagger}
%! eta* = \eta
%! eps* = \epsilon
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! bra = \bra
%! singleton = \singleton
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! -> = \rightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! Sigma = \Sigma
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! Pow = \mathcal{P}
%! Pi = \Pi
%! PiT = \Pi^{o}
%! PiEE = \Pi^{\eta\epsilon}_{+}
%! PiT = \Pi^{o}
%! PiTF = \Pi^{/}
%! bullet = \bullet
%! * = \times

@@

\title{Reversible Higher-Order Programs \\ (Functional Pearl)}
\authorinfo{Jacques Carette \and Amr Sabry}
           {McMaster University \and Indiana University}
           {carette@mcmaster.ca \and sabry@indiana.edu}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background: {{Pi}} and {{PiTF}} }
\label{sec:pi}

@@

\noindent 
Types include the unit type {{1}}, sum types {{b1+b2}}, and product types
{{b1*b2}}.  Values include {{()}} which is the only value of type {{1}},
{{left v}} and {{right v}} which inject~{{v}} into a sum type, and
{{(v1,v2)}} which builds a value of product type.  But these values should be
regarded as largely ancillary: we do not treat them as first-class citizens,
and they only occur when observing the effect of an isomorphism.

\paragraph*{Isomorphisms.} The terms of {{Pi}} witness
type isomorphisms of the form {{b <-> b}}. They consist of base
isomorphisms, as defined below, and their composition.
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}r c l@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 & <-> & b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) & <-> & (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b & <-> & b &: identr*
%swap* :&  b1 * b2 & <-> & b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) & <-> & (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 & <-> & (b1 * b3) + (b2 * b3)~ &: factor

@@

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}: 
subcode{bnf} include main
top level term, l ::= c v

\noindent
To summarize, the syntax of {{Pi}} is given as follows.

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
%subcode{bnf} include main
% value types, b ::= 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= swap+ | assocl+ | assocr+ 
%     &|& unite | uniti | swap* | assocl* | assocr* 
%     &|& dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top level term, l ::= c v
\end{definition}

@@

Operationally, the semantics of {{Pi}} is given using two mutually recursive
interpreters: one going forward and one going backwards. The use of {{sym}}
switches control from one evaluator to the other.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types}

For every value {{v : b}}, we introduce the \emph{singleton} type 
{{ singleton{v} }} whose only element is the value {{v}}.

For every value {{v : b}}, we define the \emph{pattern} 
{{ bra{v} : singleton{v}  1}} which
matches exactly {{v}} and nothing else. The type of this pattern is a function
which

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:conc}
